---
title: "Qst_Fst_Analysis"
author: "Anoob Prakash"
date: "16 April 2019"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started and formatting the data  
### Package  
```{r}
library(MCMCglmm) # for Qst analysis
library(tidyverse)
library(forestplot) # for plotting
```

### Data 
```{r}
### trait data
tdata <- read.csv("./trait_data.csv")
tdata$Trait <- runif(nrow(tdata),15,31)
str(tdata)

#rename  column family to Family due to an error MCMCglmm throws, since the package also have an argument named family
        
### Divide the dataset according to three regions [core, margin,edge]
Core <- filter(tdata, Region == "C") 
Margin <- filter(tdata, Region == "M") 
Edge <- filter(tdata, Region == "E") 

```

## Set prior    
**Setting priors for the model**   
A unifrom non informative prior was used to get the posterior distribution given the data. 
The ‘R’ in the prior stands for residual variance and G stands for the variance due to the random eﬀects. 
The prior distibution is the package is paramaterised to nu and V4. 
In this study the values are set at what is considered to be classical. 
This prior was good enough for all the models of traits studied in the three regions.  

```{r}
### set seed
set.seed(1001)

# set priors 
modPrior=list(R=list(V=1, n=0.002), 
               G=list(G1=list(V=1, n=0.002),
               G2=list(V=1, n=0.002)))
```





## Qst within each region  
### 1. Core  
```{r}

### Qst for the core

modCore1 <-  MCMCglmm(Trait ~ 1, 
                      random= ~ Family + Population,
                      family="gaussian",
                      data=Core,
                      prior=modPrior, pr=TRUE, 
                      burnin=10000 , 
                      nitt=1000000 , 
                      thin=100)
```

**Model checks**  
MCMCglmm has model checks of its own to see if the chains converged properly and if the MCMC was able
to iterate enough times to get a good read about the probability space. If the trace of the intercept is not in
the form of a fuzzy caterpillar, that would indicate there is not enough sampling at the probability space and
the model needs to be iterated more. The density of the intercept is another way of representing the varaince
components like the fuzzy caterpillar plot.
```{r}
plot(modCore1) # unimodal distribution required
```

The other model checks that are available in MCMCglmm are given below. The ‘autocorr.diag’ was intuitive
to check for autocorrelation of the model. 
However, the ‘acf()’ seems to better represent autocorrelation.
None of the models run had problems due to autocorrelation. 
The next check would be to see if the eﬀective
sample size was larger than at least a thousand. The logic behind the eﬀective size ties the no. of itterations
(nitt), burnin and thinning (thin) to see that the posterior distribution obtained is a better representative of
the data. The eﬀective size ranged from 6000-9900 for all the models and thus passed that check.  

$$EffectiveSize = (nitt − burnin)/thin$$

The ﬁnal check before inputting the values from variance co-variance matrix of the model in to the Q ST
equation was to check to see if all the factors considered for the model passed the ‘heidel.diag’ test. 
The model usually does not pass the test in cases where the amount of sampling was not high enough. 
This study has a very big sampling size, and thus it passed for all the traits under study. 
However, even if the model does not pass this test, it is not reason enough to reject the model and the posterior distribution generated. 
However, once a posterior distribution is generated after running the model, if the distribution is not uni-modal, then that would be really bad model. 
Since a mean QST value and its conﬁdence interval are required for the comparison study with F ST , 
a distibution with more than one peak is not acceptable.
However, it should be noted that the shape of the distribution just needs to be uni modal, even if its too
skewed or too ﬂat, it still is considered a good model for the trait under study. 
```{r}
### Diagnostics 
autocorr.diag(modCore1$VCV)
summary(modCore1)
effectiveSize(modCore1$Sol)
effectiveSize(modCore1$VCV) 
heidel.diag(modCore1$VCV) # needs to pass all the tests 
# effective size more than 1000
# The distribution becomes uni modal when thin = 100

### Estimate Qst  
Qst_core_fulldataset = (modCore1$VCV[,"Population"]/
                         (modCore1$VCV[,"Population"] +
                            2*(modCore1$VCV[,"Family"])))

posterior.mode(Qst_core_fulldataset) # mode for the distribution  
HPDinterval(Qst_core_fulldataset) # CI 

summary(Qst_core_fulldataset)
plot(density(Qst_core_fulldataset)) # the unimodal peak of the distribution is a good indicator

QstPlot_core_fulldataset=as.data.frame(Qst_core_fulldataset)
dens<- density(QstPlot_core_fulldataset$var1)
dens_sum = sum(density(QstPlot_core_fulldataset$var1)$y)
plot(dens$x, dens$y/dens_sum, type = "l")
```

### Qst for the margin
```{r}
modMargin1 <-  MCMCglmm(Trait ~ 1, 
                      random= ~ Family + Population,
                      family="gaussian",
                      data=slMargin,
                      prior=modPrior, 
                      pr=TRUE, 
                      burnin=10000 , 
                      nitt=1000000 , 
                      thin=100)


plot(modMargin1)


### Diagnostics 
autocorr.diag(modMargin1$VCV)
summary(modMargin1)
effectiveSize(modMargin1$Sol)
effectiveSize(modMargin1$VCV)
heidel.diag(modMargin1$VCV) 

### Estimate Qst
Qst_margin_fulldataset =(modMargin1$VCV[,"Population"]/
         (modMargin1$VCV[,"Population"] + 
            2*(modMargin1$VCV[,"Family"])))

posterior.mode(Qst_margin_fulldataset) 
HPDinterval(Qst_margin_fulldataset)

summary(Qst_margin_fulldataset)
plot(density(Qst_margin_fulldataset)) r

QstPlot_margin_fulldataset=as.data.frame(Qst_margin_fulldataset)
dens<- density(QstPlot_margin_fulldataset$var1)
dens_sum = sum(density(QstPlot_margin_fulldataset$var1)$y)
plot(dens$x, dens$y/dens_sum, type = "l")
```


### Qst for the edge
```{r}
modEdge1 <-  MCMCglmm(Height ~ 1, 
                      random= ~ Families + population,
                      family="gaussian",
                      data=slEdge,
                      prior=modPrior, pr=TRUE, burnin=10000 , nitt=1000000 , thin=100)


plot(modEdge1)


### Diagnostics
autocorr.diag(modEdge1$VCV)
summary(modEdge1)
effectiveSize(modEdge1$Sol)
effectiveSize(modEdge1$VCV)
heidel.diag(modEdge1$VCV) 


### Estimate Qst
Qst_edge_fulldataset =(modEdge1$VCV[,"Population"]/
         (modEdge1$VCV[,"Population"] + 
            2*(modEdge1$VCV[,"Family"])))

posterior.mode(Qst_edge_fulldataset) 
HPDinterval(Qst_edge_fulldataset)

summary(Qst_edge_fulldataset)
plot(density(Qst_edge_fulldataset))

QstPlot_edge_fulldataset=as.data.frame(Qst_edge_fulldataset)
dens<- density(QstPlot_edge_fulldataset$var1)
dens_sum = sum(density(QstPlot_edge_fulldataset$var1)$y)
plot(dens$x, dens$y/dens_sum, type = "l")

```


### $F_{ST}$


```{r}
fst <- read.table("Fst_distribution_regions.txt")
nrow(fst) #90000
# fst <- na.omit(fst)
# nrow(fst)

coreFst <- na.omit(fst[,1])
marginFst <- na.omit(fst[,2])
edgeFst <- na.omit(fst[,3])

```

### Fst mean and CI
```{r}
summary(coreFst) #mean = 0.01254 
quantile(coreFst,.975) # 0.088 
quantile(coreFst,.025) # 0

summary(marginFst) #mean = 0.02358 
quantile(marginFst,.975) #  0.1692 
quantile(marginFst,.025) # 0

summary(edgeFst) #mean = 0.01892  
quantile(edgeFst,.975) #  0.1394 
quantile(edgeFst,.025) # 0
```


### Bootstrapping function
```{r}
#####################bootstrapping

# a function to perform bootstrapping
boot.mean.sampling.distribution = function(raw.data, B=1000){
  # this function will take 1,000 (by default) bootsamples calculate the mean of 
  # each one, store it, & return the bootstrapped sampling distribution of the mean

  boot.dist = vector(length=B)     # this will store the means
  N         = length(raw.data)     # this is the N from your data
  for(i in 1:B){
    boot.sample  = sample(x=raw.data, size=N, replace=TRUE)
    boot.dist[i] = mean(boot.sample)
  }
  boot.dist = sort(boot.dist)
  return(boot.dist)
}
```

### Bootstrapping Fst for each region
```{r}
coreB <- boot.mean.sampling.distribution(coreFst)
mean(coreB) # 0.01253692
hist(coreB)
sd(coreB)

quantile(coreB,.025) # lwr = 0.01236278 
quantile(coreB,.975) # upr = 0.01271652 

# ------------------------------------------------------------


marginB <- boot.mean.sampling.distribution(marginFst)

mean(marginB) # 0.02357289
hist(marginB)
sd(marginB)

quantile(marginB,.025) # lwr = 0.02321597
quantile(marginB,.975) # upr = 0.02394149 

# ------------------------------------------------------------

edgeB <- boot.mean.sampling.distribution(edgeFst)

mean(edgeB) # 0.01892543
hist(edgeB)
sd(edgeB)

quantile(edgeB,.025) # lwr = 0.01860426
quantile(edgeB,.975) # upr = 0.01922697 
```

### Plotting Qst vs Fst
```{r}
# library(forestplot)

coreQst_Fst <- na.omit(read.table("coreQst_Fst.csv", sep = ",", header = TRUE))



tabletext<-cbind(
  c("", "Height", "Seed_weight", "Relative_fitness", "Core Fst"))

xticks <- seq(from = -.025, to = 1, by = 0.25)
xtlab <- rep(c(TRUE, FALSE), length.out = length(xticks))
attr(xticks, "labels") <- xtlab

clrs <- fpColors(box="darkred",line="darkblue", summary="royalblue")

coreMeta <- structure(list(
    mean  = c(NA, 0.000127193,0.145735000, 0.001161110, 0.0125369200), 
    lower = c(NA, 0.000002577,0.058135200, 0.000036917, 0.0123627800),
    upper = c(NA, 0.139271600,0.347303800, 0.174457100, 0.0127165200)),
    .Names = c("mean", "lower", "upper"), 
    row.names = c(NA, -4L), 
    class = "data.frame")


forestplot(tabletext,
           coreMeta,new_page = TRUE,
           clip=c(0.001,1.0),
           xlog=TRUE,
           boxsize = .10,  # We set the box size to better visualize the type
           line.margin = .1, # We need to add this to avoid crowding
           col=clrs
           
           )

## Core plot

clrs <- fpColors(box="darkred",line="darkblue", summary="royalblue")
xticks <- seq(from = -.0, to = 1.15, by = 0.25)
forestplot(tabletext,
           coreMeta,new_page = TRUE,
           clip=c(-.025, 1),
           boxsize = .10,  # We set the box size to better visualize the type
           line.margin = .1, # We need to add this to avoid crowding
           col=clrs,
           xticks = xticks
           )

clrs <- fpColors(box="darkred",line="darkblue", circle="royalblue")
xticks <- seq(from = -.0, to = 1.15, by = 0.25)
forestplot(tabletext,
           coreMeta,new_page = TRUE,
           fn.ci_norm = c(fpDrawNormalCI,fpDrawNormalCI, fpDrawNormalCI, fpDrawCircleCI),
           clip=c(-.025, 1),
           boxsize = .10,  # We set the box size to better visualize the type
           line.margin = .1, # We need to add this to avoid crowding
           col=clrs,
           xticks = xticks
           )


# Forest plot using ggplot
p = ggplot(data=RR_data,
    aes(x = Group,y = RiskRatio, ymin = LowerLimit, ymax = UpperLimit ))+
    geom_pointrange(aes(col=Group))+
    geom_hline(aes(fill=Group),yintercept =1, linetype=2)+
    xlab('Group')+ ylab("Risk Ratio (95% Confidence Interval)")+
    geom_errorbar(aes(ymin=LowerLimit, ymax=UpperLimit,col=Group),width=0.5,cex=1)+ 
    facet_wrap(~Condition,strip.position="left",nrow=9,scales = "free_y") +
    theme(plot.title=element_text(size=16,face="bold"),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_text(face="bold"),
        axis.title=element_text(size=12,face="bold"),
        strip.text.y = element_text(hjust=0,vjust = 1,angle=180,face="bold"))+
    coord_flip()
```

### Qst/Fst

```{r}
### Core Qst/Fst plot

library(readxl)
coreXL <- data.frame(read_excel("coreQst_Fst.xlsx"))

corePlot <- ggplot(data=coreXL,
    aes(x = Type,y = Mean, ymin = LowerLimit, ymax = UpperLimit ))+
    geom_pointrange(aes(col=Type))+
    geom_hline(aes(fill=Type),yintercept = 0.012536920, linetype=2)+
    xlab('')+ ylab("Qst/Fst")+
    geom_errorbar(aes(ymin=LowerLimit, ymax=UpperLimit,col=Type),width=0.5,cex=1)+ 
    facet_wrap(~Value,strip.position="left",nrow=9,scales = "free_y") +
    theme(plot.title=element_text(size=24,face="bold"),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_text(face="bold"),
        axis.title=element_text(size=14,face="bold"),
        strip.text.y = element_text(size = 14, hjust=0.5,vjust = 0.5,angle=180,face="bold"))+
    coord_flip()

corePlot + theme_classic()

corePlot + theme_bw()

corePlot + ylim(0, 1)


### Margin Qst/Fst plot
marginXL <- data.frame(read_excel("marginQst_Fst.xlsx"))

marginPlot <- ggplot(data=marginXL,
    aes(x = Type,y = Mean, ymin = LowerLimit, ymax = UpperLimit ))+
    geom_pointrange(aes(col=Type))+
    geom_hline(aes(fill=Type),yintercept = 0.0235728900, linetype=2)+
    xlab('')+ ylab("Qst/Fst")+
    geom_errorbar(aes(ymin=LowerLimit, ymax=UpperLimit,col=Type),width=0.5,cex=1)+ 
    facet_wrap(~Value,strip.position="left",nrow=9,scales = "free_y") +
    theme(plot.title=element_text(size=24,face="bold"),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_text(face="bold"),
        axis.title=element_text(size=14,face="bold"),
        strip.text.y = element_text(size = 14, hjust=0.5,vjust = 0.5,angle=180,face="bold"))+
    coord_flip()

marginPlot + theme_classic()

marginPlot + theme_bw()

marginPlot +  ylim(0, 1)





### Edge Qst/Fst plot
edgeXL <- data.frame(read_excel("edgeQst_Fst.xlsx"))

edgePlot <- ggplot(data=edgeXL,
    aes(x = Type,y = Mean, ymin = LowerLimit, ymax = UpperLimit ))+
    geom_pointrange(aes(col=Type))+
    geom_hline(aes(fill=Type),yintercept = 1.892543e-02, linetype=2)+
    xlab('')+ ylab("Qst/Fst")+
    geom_errorbar(aes(ymin=LowerLimit, ymax=UpperLimit,col=Type),width=0.5,cex=1)+ 
    facet_wrap(~Value,strip.position="left",nrow=9,scales = "free_y") +
    theme(plot.title=element_text(size=24,face="bold"),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_text(face="bold"),
        axis.title=element_text(size=14,face="bold"),
        strip.text.y = element_text(size = 14, hjust=0.5,vjust = 0.5,angle=180,face="bold"))+
    coord_flip()

edgePlot + theme_classic() + ylim(0, 1)

edgePlot + theme_bw()

edgePlot  + ylim(0, 1)

edgePlot  + ylim(0, 1) + theme_bw()




```

